Active-contours driven registration method for structure-informed segmentation of diffusion MR images

### 2015-04-16 Publications/2015-NeuroImage/03-introduction.tex
version sent to proofreading

Early attempts to delineate the \gls*{wm} proceeded by thresholding the 
  \gls*{fa}\footnote{the \gls*{fa} is a scalar parameter of diffusion derived from
  the \gls*{dmri} data.} map.
However, the mask and subsequent analyses substantially depend on the threshold set
  \citep{taoka_fractional_2009}.
To overcome the unreliability of of \gls*{fa} thresholding,
  \cite{zhukov_level_2003} proposed active contours with edges represented
  by levelsets, evolving on directionally invariant scalar maps.
\cite{rousson_level_2004} successfully segmented the corpus callosum with
  region-based levelsets on the eigenvalues of tensors fitted on the
  \gls*{dmri} data.
A large body of work targeted the definition of appropriate features to cluster,
  such as the 5D representation proposed by \cite{jonasson_segmentation_2005}.
Other efforts include mixture models on sets of directionally invariant maps
  \citep{liu_brain_2007}, iterative \citep{hadjiprocopis_unbiased_2005} and
  hierarchical \citep{lu_segmentation_2008} clusterings,
  graph-cuts \citep{han_experimental_2009},
  and volume fraction modeling \citep{kumazawa_improvement_2013}.

To solve the segmentation problem through registration, \cite{saad_new_2009} 
  used Pearson correlation to perform linear alignment of the \gls*{t1} and 
  the \emph{b0}.
Similarly, an also linear-registration method is \emph{bbregister} \citep{greve_accurate_2009}.
It uses active contours with edges to look for intensity boundaries in the \emph{b0}
  image.
The active contours are initialized by surfaces extracted from the 
  \gls*{t1} using \emph{FreeSurfer} \citep{fischl_freesurfer_2012}:
  the \emph{pial} surface (exterior of cortical \gls*{gm}) and the \emph{white}
  surface (the \gls*{wm}/\gls*{gm} interface). 
Since the \emph{b0} image only presents a detectable frontier for the pial surface,
  \emph{bbregister} is limited to align the cortical layer in this
  application.
This tool has become the standard method for its proven robustness, despite the 
  distortions found in \gls*{dmri} are nonlinear.
To overcome this issue, \emph{bbregister} excludes from the boundary search those
  regions typically warped by the artifact.
As the distortion is not accounted for, it must be addressed separately.
Nonlinear registration has been successfully performed between \gls*{t2} and \emph{b0}
  images for their similarity, but uniquely as a way to correct for distortions
  \citep{kybic_unwarping_2000,studholme_accurate_2000,wu_comparison_2008,tao_variational_2009}.
Further registration of \gls*{t1} and \gls*{t2} images was still required to map the anatomical
  information (and the \gls*{wm} segmentation) into \gls*{dmri} space.

Therefore, given the described circumstances, a method performing simultaneously
  segmentation in native \gls*{dmri} space and registration of the corresponding \gls*{t1} image
  to the \gls*{dmri} data could provide the optimal solution.
As far as we know, such strategy has never been proposed for the application at hand.
Previously, joint segmentation and registration has been successfully applied to other problems.
For instance, longitudinal object tracking \citep{paragios_level_2003} and atlas-based
  segmentation \citep{gorthi_active_2011}.
The most extended approach is integrating active contours in image registration 
  methods.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  improved an existing method \citep{yezzi_variational_2003} based on linear registration
  to the nonlinear case implementing a free-form deformation field.
\cite{droske_mumfordshah_2009} reviewed the existing techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration and segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method in 2D using levelsets and a nonlinear elasticity smoother on the
  displacement vector field, which preserves topology even with very large deformations.
Finally, \cite{gorthi_active_2011} extended the existing methodologies using a multiphase
  levelset function for the registration of several active contours, in the application
  of atlas-based segmentation.

>>>>

Early attempts to delineate the \gls*{wm} proceeded by thresholding the 
  \gls*{fa}\footnote{the \gls*{fa} is a scalar parameter of diffusion derived from
  the \gls*{dmri} data.} map.
However, the mask and subsequent analyses substantially depend on the threshold set
  \citep{taoka_fractional_2009}.
To overcome the unreliability of \gls*{fa} thresholding,
  \cite{zhukov_level_2003} proposed active contours with edges represented
  by levelsets, evolving on directionally invariant scalar maps.
\cite{rousson_level_2004} successfully segmented the corpus callosum with
  region-based levelsets on the eigenvalues of tensors fitted on the
  \gls*{dmri} data.
A large body of work targeted the definition of appropriate features to cluster,
  such as the 5D representation proposed by \cite{jonasson_segmentation_2005}.
Other efforts include mixture models on sets of directionally invariant maps
  \citep{liu_brain_2007}, iterative \citep{hadjiprocopis_unbiased_2005} and
  hierarchical \citep{lu_segmentation_2008} clusterings,
  graph-cuts \citep{han_experimental_2009},
  and volume fraction modeling \citep{kumazawa_improvement_2013}.

To solve the segmentation problem through registration, \cite{saad_new_2009} 
  used Pearson correlation to perform a linear alignment of the \gls*{t1} and 
  the \emph{b0}.
Similarly, an also linear-registration method is \emph{bbregister} \citep{greve_accurate_2009}.
It uses active contours with edges to look for intensity boundaries in the \emph{b0}
  image.
The active contours are initialized by surfaces extracted from the 
  \gls*{t1} using \emph{FreeSurfer} \citep{fischl_freesurfer_2012}:
  the \emph{pial} surface (exterior of cortical \gls*{gm}) and the \emph{white}
  surface (the \gls*{wm}/\gls*{gm} interface). 
Since the \emph{b0} image only presents a detectable frontier for the pial surface,
  \emph{bbregister} is limited to align the cortical layer in this
  application.
This tool has become the standard method for its proven robustness, despite the 
  distortions found in \gls*{dmri} are nonlinear.
To circumvent this issue, \emph{bbregister} excludes from the boundary search those
  regions typically warped by the artifact.
As the distortion is not accounted for, it must be addressed separately.
Nonlinear registration has been successfully performed between \gls*{t2} and \emph{b0}
  images for their similarity, but uniquely as a way to correct for distortions
  \citep{kybic_unwarping_2000,studholme_accurate_2000,wu_comparison_2008,tao_variational_2009}.
Further registration of \gls*{t1} and \gls*{t2} images was still required to map the anatomical
  information (and the \gls*{wm} segmentation) into \gls*{dmri} space.

Therefore, given the described circumstances, a method performing simultaneously
  segmentation in native \gls*{dmri} space and registration of the corresponding \gls*{t1} image
  to the \gls*{dmri} data could provide the optimal solution.
As far as we know, such strategy has never been proposed for the application at hand.
Previously, joint segmentation and registration has been successfully applied to other problems.
For instance, longitudinal object tracking \citep{paragios_level_2003} and atlas-based
  segmentation \citep{gorthi_active_2011}.
The most extended approach is integrating active contours in image registration 
  methods.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  improved an existing method \citep{yezzi_variational_2003} based on linear registration
  to the nonlinear case implementing a free-form deformation field.
\cite{droske_mumfordshah_2009} reviewed the existing techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration and segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method in 2D using levelsets and a nonlinear elasticity smoother on the
  displacement vector field, which preserves the topology even with very large deformations.
Finally, \cite{gorthi_active_2011} extended the existing methodologies using a multiphase
  levelset function for the registration of several active contours, in the application
  of atlas-based segmentation.

### 2015-04-16 Publications/2015-NeuroImage/04-methods.tex
version sent to proofreading

\begin{figure*}
\includegraphics[width=\linewidth]{figure02}
\caption{The evaluation of \regseg{} on phantom data is performed with the presented instrumental workflow:
  1) The reference surfaces $\gammaset_R$ are triangularized meshes extracted from the four binary shapes (``box'', ``ball'', ``L'', ``gyrus'');
  2) A ground-truth displacement field is generated as described in \autoref{sec:digital_phantoms}, and applied to warp
      $\gammaset_R$, obtaining $\gammaset_\text{true}$;
  3) Once warped, $\gammaset_\text{true}$ are projected to the corresponding discrete 3D volume, and downsampled creating partial volume effects to two resolutions
     of \isores[mm]{2.0} and \isores[mm]{1.0}, producing sets of tissue fractions maps;
  4) The tissue fractions are fed to an \acrfull*{mr} simulator, generating \acrfull*{t1} and \acrfull*{t2} -like images at the
     two available resolutions;
  5) The \regseg{} is run, using the warped test images as multispectral moving-image and $\gammaset_R$ as shape-priors;
  6) Assessment of the agreement between the surfaces fitted with \regseg{} ($\hat{\gammaset}_{test}$) and $\gammaset_\text{true}$
     with automatically generated visual reports and computing the Hausdorff distance between corresponding surfaces.}\label{fig:evphantoms}
\end{figure*}

@@

\begin{figure*}
\includegraphics[width=\linewidth]{figure03}
\caption{Experimental workflow real data from the \acrfull*{hcp}.
  1) $\gammaset_R$ are extracted from the anatomical reference (\gls*{t1} image).
  2) To operate as ground truth, we generate a plausible-synthetic distortion $U_{true}$
    from the fieldmap using \eqref{eq:fieldmap}.
  3) The \gls*{dmri} data are warped using $U_{true}$ to reproduce the effects of real
    susceptibility-derived distortions.
  Target diffusion scalars (\gls*{fa} and \gls*{adc}) are computed on the distorted data and
    stacked to feed the multivariate input required by \regseg{}.
  4) the method is run, obtaining a $U_{test} = \hat{U}_{true}$, the estimation of
    the ground-truth deformation.
  5) Results are visually and quantitatively evaluated.}\label{fig:evworkflows}
\end{figure*}

>>>>

\begin{figure*}
\includegraphics[width=\linewidth]{figure02}
\caption{The evaluation of \regseg{} on phantom data is performed with the presented instrumental workflow:
  1) The reference surfaces $\gammaset_R$ are triangularized meshes extracted from the four binary shapes (``box'', ``ball'', ``L'', ``gyrus'');
  2) A ground-truth displacement field is generated as described in \autoref{sec:digital_phantoms}, and applied to warp
      $\gammaset_R$, obtaining $\gammaset_\text{true}$;
  3) Once warped, $\gammaset_\text{true}$ are projected to the corresponding discrete 3D volume, and downsampled creating partial volume effects to two resolutions
     of \isores[mm]{2.0} and \isores[mm]{1.0}, producing sets of tissue fractions maps;
  4) The tissue fractions are fed to an \acrfull*{mr} simulator, generating \acrfull*{t1} and \acrfull*{t2} -like images at the
     two available resolutions;
  5) The \regseg{} tool is run, using the warped test images as multispectral moving-image and $\gammaset_R$ as shape-priors;
  6) Assessment of the agreement between the surfaces fitted with \regseg{} ($\hat{\gammaset}_{test}$) and $\gammaset_\text{true}$
     with automatically generated visual reports and computing the Hausdorff distance between corresponding surfaces.}\label{fig:evphantoms}
\end{figure*}

@@

\begin{figure*}
\includegraphics[width=\linewidth]{figure03}
\caption{Experimental workflow to process real data from the \acrfull*{hcp}.
  1) $\gammaset_R$ are extracted from the anatomical reference (\gls*{t1} image).
  2) To operate as ground truth, we generate a plausible-synthetic distortion $U_{true}$
    from the fieldmap using \eqref{eq:fieldmap}.
  3) The \gls*{dmri} data are warped using $U_{true}$ to reproduce the effects of real
    susceptibility-derived distortions.
  Target diffusion scalars (\gls*{fa} and \gls*{adc}) are computed on the distorted data and
    stacked to feed the multivariate input required by \regseg{}.
  4) the method is run, obtaining a $U_{test} = \hat{U}_{true}$, the estimation of
    the ground-truth deformation.
  5) Results are visually and quantitatively evaluated.}\label{fig:evworkflows}
\end{figure*}

### 2015-04-16 Publications/2015-NeuroImage/05-results.tex
version sent to proofreading

\begin{figure*}
  \centering
  \includestandalone[width=\linewidth]{figure05}
  \caption{A. Visual assessment of the results on the low resolution sets:
    ``gyrus'' (top-left), ``L'' (top-right), ``ball'' (bottom-left),
    and ``box'' at (bottom-right).
  In yellow color, the recovered contours after registration are represented.
  Our method showed high accuracy, as it demonstrates the almost exact location of the registered
    contours with respect to their ground truth position depicted in green.
  Partial volume effect turns segmentation of the sulci a challenging problem with voxel-wise
    clustering methods, but it is successfully segmented with our method.
  B. Quantitative evaluation of registration error in terms of average Hausdorff distance of
    surfaces at low (left) and high (right) resolutions, demonstrating that the error is
    consistently below the voxel size.
    }\label{fig:phantom}
\end{figure*}
A total of 1200 experiments (4 phantom types $\times$ 150 random warpings $\times$ 2 resolutions) were
  run using the workflow given in \autoref{fig:evphantoms}.
For each experiment, the misregistration error was measured using the Hausdorff distance
  (see \autoref{sec:experiments_evaluation}) between the theoretical $\gammaset_\text{true}$ and
  the estimation done by \regseg{} ($\hat{\gammaset}_{test}$).
The results showed a consistent and high accuracy, below the image resolution.
\autoref{fig:phantom} (block C) presents the violin plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
In order to relate average misregistration error to the resolution of the moving image,
  we proceeded as follows.
First, we confirmed that the vertex-wise error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$, for each phantom type).
Averaged errors resulted significantly lower than voxel size with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals,
  reported in \autoref{tab:ci_phantom}.

@@

\subsection{Evaluation on real datasets and cross-comparison}\label{sec:results_hcp}
%
Finally, we compared the performance of \regseg{} and the standard \gls*{t2b}
  method.
Visual assessment of the 16 cases is included in the \suppl{section S5}.
In \autoref{fig:results_real}, box A, the visual report for one subject is supplied.
We computed the \gls*{swindex} \eqref{eq:swindex} of every surface after registration,
  for both \regseg{} and the \gls*{t2b} methods.
Finally, to statistically compare the results, we performed Kruskal-Wallis H-tests
  (a non-parametric alternative to ANOVA) on the warping indices for the three surfaces of 
  interest selected in \autoref{sec:experiments_evaluation}
  ($\Gamma_{VdGM}$, $\Gamma_{WM}$, $\Gamma_{pial}$).
All the statistical tests evidenced that error distributions obtained with \regseg{} and
  \gls*{t2b} were significantly different, and using the violin plots in box B of
  \autoref{fig:results_real} we observed that errors are always larger for \gls*{t2b}.
We also reported the 95\% CIs of the \gls*{swindex} for those surfaces.
The aggregate CI for \regseg{} was 1.08 - 1.50 [mm], whereas the \gls*{t2b} method
  yielded an aggregate CI of 2.06 - 2.43 [mm].
The results of the statistical tests and the CIs are summarized in \autoref{tab:results_real}.

>>>>

\begin{figure*}
  \centering
  \includestandalone[width=\linewidth]{figure05}
  \caption{A. Visual assessment of the results on the low resolution sets:
    ``gyrus'' (top-left), ``L'' (top-right), ``ball'' (bottom-left),
    and ``box'' at (bottom-right).
  In yellow color, the recovered contours after registration are represented.
  \Regseg showed high accuracy, as it demonstrates the almost exact location of the registered
    contours with respect to their ground truth position depicted in green.
  Partial volume effect turns segmentation of the sulci a challenging problem with voxel-wise
    clustering methods, but it is successfully segmented with \regseg{}.
  B. Quantitative evaluation of registration error in terms of average Hausdorff distance of
    surfaces at low (left) and high (right) resolutions, demonstrating that the error is
    consistently below the voxel size.
    }\label{fig:phantom}
\end{figure*}
A total of 1200 experiments (4 phantom types $\times$ 150 random warpings $\times$ 2 resolutions) were
  run using the workflow given in \autoref{fig:evphantoms}.
For each experiment, the misregistration error was measured using the Hausdorff distance
  (see \autoref{sec:experiments_evaluation}) between the theoretical $\gammaset_\text{true}$ and
  the estimation done by \regseg{} ($\hat{\gammaset}_{test}$).
The results showed a consistent and high accuracy, below the image resolution.
\autoref{fig:phantom} (block C) presents the violin plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
In order to relate average misregistration error to the resolution of the moving image,
  we proceeded as follows.
First, we confirmed that the vertex-wise error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$, for each phantom type).
Averaged errors resulted significantly lower than voxel size with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals,
  reported in \autoref{tab:ci_phantom}.

@@

\subsection{Evaluation on real datasets and cross-comparison}\label{sec:results_hcp}
%
Finally, we compared the performance of \regseg{} and the standard \gls*{t2b}
  method.
Visual assessment of the 16 cases is included in the \suppl{section S5}.
In \autoref{fig:results_real}, box A, the visual report for one subject is supplied.
We computed the \gls*{swindex} \eqref{eq:swindex} of every surface after registration,
  for both \regseg{} and the \gls*{t2b} methods.
Finally, to statistically compare the results, we performed Kruskal-Wallis H-tests
  (a non-parametric alternative to ANOVA) on the warping indices for the three surfaces of 
  interest selected in \autoref{sec:experiments_evaluation}
  ($\Gamma_{VdGM}$, $\Gamma_{WM}$, $\Gamma_{pial}$).
All the statistical tests evidenced that error distributions obtained with \regseg{} and
  \gls*{t2b} were significantly different, and using the violin plots in box B of
  \autoref{fig:results_real} we observed that errors were always larger for \gls*{t2b}.
We also reported the 95\% CIs of the \gls*{swindex} for those surfaces.
The aggregate CI for \regseg{} was 1.08 - 1.50 [mm], whereas the \gls*{t2b} method
  yielded an aggregate CI of 2.06 - 2.43 [mm].
The results of the statistical tests and the CIs are summarized in \autoref{tab:results_real}.

### 2015-04-16 Publications/2015-NeuroImage/06-discussion.tex
version sent to proofreading

\paragraph*{Accuracy tests}
The hypothesis underpinning our research was that image registration can be reliably performed
  by searching for homogeneous regions in a multispectral image that correspond to precise contours
  from an atlas, or extracted from other image (i.e. a different time step).
We proved that active contours without edges can be used to successfully drive a
  deformation supported by B-Spline basis functions using digital phantoms.
We randomly deformed four different phantom models mimicking three homogeneous regions,
  and use them to simulate \gls*{t1} and \gls*{t2} images at two resolution levels.
After registration with \regseg{} we measured the Hausdorff distance between the
  projected contours using the ground-truth warping and our estimation.
We concluded that errors are significantly lower than voxel resolution.
We also assessed the 95\% \gls*{ci}, yielding an aggregate interval of
  0.64 - 0.66 [mm] in the low resolution phantoms (2.0mm isotropic voxel) and
  0.34 - 0.38 [mm] in the high resolution phantoms (1.0mm isotropic).
Therefore, we also concluded that the error was bounded above by the half of the
  voxel spacing.

>>>>

\paragraph*{Accuracy tests}
The hypothesis underpinning our research was that image registration can be reliably performed
  by searching for homogeneous regions in a multispectral image that correspond to precise contours
  from an atlas, or extracted from other image (i.e. a different time step).
We proved that active contours without edges can be used to successfully drive a
  deformation supported by B-Spline basis functions using digital phantoms.
We randomly deformed four different phantom models mimicking three homogeneous regions
  (\gls*{wm}, \gls*{gm}, and \acrlong*{csf}), and use them to simulate \gls*{t1} and \gls*{t2}
  images at two resolution levels.
After registration with \regseg{}, we measured the Hausdorff distance between the
  projected contours using the ground-truth warping and our estimation.
We concluded that errors are significantly lower than voxel resolution.
We also assessed the 95\% \gls*{ci}, yielding an aggregate interval of
  0.64 - 0.66 [mm] in the low resolution phantoms (2.0mm isotropic voxel) and
  0.34 - 0.38 [mm] in the high resolution phantoms (1.0mm isotropic).
Therefore, we also concluded that the error was bounded above by the half of the
  voxel spacing.
